Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-10 paper/main.tex
definitions and pseudocode

\begin{abstract}
	Numerous causal discovery algorithms were developed to automatically learn
	directed acyclic graphs (DAGs) and other causal models from data. However,
	their adoption in applied domains remains limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to produce results that contradict
	obvious domain knowledge and their inability to distinguish Markov equivalent
	models. To assist researchers in constructing DAGs manually, we propose an iterative
	structure learning approach that combines domain knowledge with
	data-driven insights. Our method leverages conditional independence
	testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	In a simulation study, we find that this iterative approach to leverage domain 
	knowledge already starts outperforming purely data-driven structure learning if 
	the orientation of new edge is correctly determined in at least two out of three cases.
	We present a proof-of-concept implementation using a large language 
	model as a domain expert and a graphical user interface designed to 
	assist human experts with DAG construction.
\end{abstract}

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. In the DAG literature, the primary
focus has been on developing automated algorithms to learn causal structures
from datasets. These efforts have led to numerous causal discovery algorithms,
such as constraint-based methods like PC algorithm \citep{Spirtes2001,KalischB07} 
and Fast Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NOTEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite significant progress in automated causal discovery,
their adoption in applied research has been limited. Challenges encountered with existing
causal discovery algorithms in practice include but are not limited to:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While constraint-based algorithms are 
		often asymptotically
		consistent \citep{KalischB07}, they can and do make mistakes
		on  finite samples. These mistakes can be severe and contradict
		obvious domain knowledge (think of edges going into unmodifiable
		attributes such as Age). The choice of algorithm and hyperparameters
		significantly affects the output, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. A recent paper
		therefore advised Epidemiologists to not attempt using structure
		learning algorithms without help of an expert \citep{Gururaghavendran_2024}.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. This structural uncertainty
		can make it difficult or impossible to apply the learned model for downstream
		tasks, such as identification or causal
		effect estimation \citep{Maathuis_2009,PerkovicTKM17}.
\end{enumerate}

@@

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the standard
deviation of $ X $ as $ \sigma_X $, the covariance between variables $ X $ and
$ Y $ as $ \mathrm{cov}(X, Y) $, a covariance matrix as $ \Sigma $, and the
entry corresponding to the covariance between $ X $ and $ Y $ as $ \Sigma_{XY}
$. We consider causal DAGs that contain only observed variables. A DAG $ G =
(V, E) $ is an acyclic directed graph whose nodes $ V $ corresponding to random
variables and whose edges $ E $ represent direct causal relationships
\cite{Pearl2009}. For an edge $e=X \to Y$, $\textrm{reverse}(e)=Y \to X$. The
set of parents of $ X $ in $ G $ is denoted as $ \textrm{Pa}_G(X) $, and its
ancestors and descendants as $ \textrm{An}_G(X) $ and $ \textrm{De}_G(X) $,
respectively; we use the convention that $X \notin \textrm{An}_G(X)$ and $X
\notin \textrm{De}_G(X)$. We denote a path, $ \pi(X, Y) = \{ X, V_0, V_1,
\cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G $ where consecutive pairs of
variables in $ \pi $ are connected by an edge. We say $ X $ and $ Y $ are
d-connected in $ G $ given a conditioning set $ \bm{Z} \subseteq V - \{X, Y\}
$, if for a path $ \pi(X, Y) $ (i) for every collider structure ($ V_1
\rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $ \bm{Z} \cap \{ V_2,
\textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every non-collider
structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1 \leftarrow V_2
\rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are d-separated
given $ \bm{Z} $ if no d-connecting path exist between them.

@@

We consider two models of a domain experts. Given two variables
$ X $ and $ Y $ that are part of a directed graph $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$
Note that the ancestral oracle does not consider differences between direct and 
indirect relationships: for any $G, H$ where $G^+=H^+$, we have $\mathcal{A}_G=\mathcal{A}_H$.

@@

These two oracles still provide correct information on all ancestral relationships that are 
present in $G$, but cannot reliably detect the \emph{absence} of ancestral relationships.
Further, the inconsistent ancestral oracle sometimes contradicts itself: there are at least two
variables $X,Y$ where its information implies both a causal effect of $X$ on $Y$ and a reverse 
causal effect of $Y$ on $X$. Therefore, if we use such oracles for DAG construction, we will
introduce superfluous edges into our graph.

Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$, and $0$ otherwise. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \cite{Spirtes2001}.

Based on these two oracles, we now define two core procedures that
iteratively change a current DAG structure. The procedure 
\textsc{Expand} (Algorithm~\ref{algo:expand}) uses CI  
information (supplied in this case by the d-separation oracle) 
to search for unexplained associations in the graph and 
then uses domain knowledge to determine ancestry. The 
procedure \textsc{Prune} (Algorithm~\ref{algo:prune}) uses
CI information to remove superfluous edges from the
graph.

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    $L \gets L \cup \{ e \} $ \;
	    \If{ $ (V, E \cup L) $ is cyclic }{
		    $ L \gets \textsc{FixCycles}(V, E, L, e, \mathcal{D}_G) $ \;
	    }
	    % \If{ $(V,E \cup L \cup e)$ is acyclic }{
            % 	$L \gets L \cup  \{ e \} $\;
	    % } \Else {
            % 	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    % }
        }
	\If{$|L|\geq k$}{\KwGoTo 16\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycle}{FixCycle}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, L, e, \mathcal{D}_G $)}{
	$ u, v \gets e $ \;
	\ForEach {$ P = \{ v \rightarrow x_1, x_1 \rightarrow x_2, \cdots,  x_{k-1} \rightarrow x_k, x_k \rightarrow u \} $ in ($V, E \cup L$) }{
		\ForEach {$ x_i \rightarrow x_j $ in $ P $}{
        	let $\mathbf{Z}$ be a set that $d$-separates $x_i$ and $x_j$ in $(V, E \cup L)$\;
		\If {$ \mathcal{D}_G(X, Y, \mathbf{Z}) = 0 $}{
			$ L \gets L - \{ x_i \rightarrow x_j \} $ \;

		}
	}
	}
	\Return $ L $ \;
}
\caption{Fixing cycles by removing incorrect edges.}
\label{algo:fix_cycles}
\end{algorithm}

>>>>

\begin{abstract}
	Many researchers construct directed acyclic graph (DAG) models manually
	based on domain knowledge. 
	Although numerous causal discovery algorithms were developed to automatically learn
	DAGs and other causal models from data, these remain challenging to 
	use due to their tendency to produce results that contradict
	domain knowledge, among other issues. Here we propose a hybrid, 
	iterative structure learning approach that combines domain knowledge with
	data-driven insights to assist researches in constructing DAGs.
	Our method leverages conditional independence
	testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	In a simulation study, we find that this iterative approach to leverage domain 
	knowledge already starts outperforming purely data-driven structure learning if 
	the orientation of new edge is correctly determined in at least two out of three cases.
	We present a proof-of-concept implementation using a large language 
	model as a domain expert and a graphical user interface designed to 
	assist human experts with DAG construction.
\todo[inline]{Should we add links here to the implementations? Is the two out of three cases still correct?}
\end{abstract}

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. These
include constraint-based methods like PC algorithm \citep{Spirtes2001,KalischB07} 
and Fast Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NOTEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite this significant body of work,
the adoption of causal discovery methods in observational research has so far 
been limited. Challenges encountered with existing
causal discovery algorithms in practice include but are not limited to:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While constraint-based algorithms are 
		often asymptotically
		consistent \citep{KalischB07}, they can and do make mistakes
		on  finite samples. These mistakes can be severe and contradict
		obvious domain knowledge (think of edges going into unmodifiable
		attributes such as Age). The choice of algorithm and hyperparameters
		significantly affects the output, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. A recent paper
		 advised Epidemiologists to not attempt using structure
		learning algorithms without help of an expert \citep{Gururaghavendran_2024}.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. This structural uncertainty
		can make it difficult or impossible to apply the learned model for downstream
		tasks, such as identification or causal
		effect estimation \citep{Maathuis_2009,PerkovicTKM17}.
\end{enumerate}

@@

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the standard
deviation of $ X $ as $ \sigma_X $, the covariance between variables $ X $ and
$ Y $ as $ \mathrm{cov}(X, Y) $, a covariance matrix as $ \Sigma $, and the
entry corresponding to the covariance between $ X $ and $ Y $ as $ \Sigma_{XY}
$. A DAG $ G =
(V, E) $ is an acyclic directed graph whose nodes $ V $ corresponding to random
variables and whose edges $ E $ represent direct causal relationships
\cite{Pearl2009}. For an edge $e=X \to Y$, $\textrm{reverse}(e)=Y \to X$. The
set of parents of $ X $ in $ G $ is denoted as $ \textrm{Pa}_G(X) $, and its
ancestors and descendants as $ \textrm{An}_G(X) $ and $ \textrm{De}_G(X) $,
respectively; we use the convention that $X \notin \textrm{An}_G(X)$ and $X
\notin \textrm{De}_G(X)$. We denote a path, $ \pi(X, Y) = \{ X, V_0, V_1,
\cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G $ where consecutive pairs of
variables in $ \pi $ are connected by an edge. We say $ X $ and $ Y $ are
d-connected in $ G $ given a conditioning set $ \bm{Z} \subseteq V - \{X, Y\}
$, if for a path $ \pi(X, Y) $ (i) for every collider structure ($ V_1
\rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $ \bm{Z} \cap \{ V_2,
\textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every non-collider
structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1 \leftarrow V_2
\rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are d-separated
given $ \bm{Z} $ if no d-connecting path exist between them.

\subsection{Assumptions} 
In this paper, we consider the structure learning problem under the widely
known \emph{causal Markov}, \emph{causal sufficiency} and \emph{faithfulness}
assumptions \citep{Spirtes1993}. These entail that there exists a 
DAG $G$ on the variables $\bm{X}$ whose implied $d$-separation statements coincide
exactly with the conditional independence relationships among the variables 
in $\bm{X}$, and for which all direct common causes of variables in $\bm{X}$ are 
also included in $\bm{X}$. These are the same assumptions made by the structure
learning algorithms we consider in this paper, such as the PC algorithm \citep{Spirtes1993}.

@@

We model domain experts as procedures that take two variables $X$ and $Y$ that are assumed 
to be part of a DAG $G$ and 
provide information on the ancestral relationship between them. First, 
a \emph{strong ancestral oracle} $\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$
Note that the ancestral oracle does not consider differences between direct and 
indirect relationships: for any $G, H$ where $G^+=H^+$, we have $\mathcal{A}_G=\mathcal{A}_H$.

@@

These two oracles still provide correct information on all ancestral relationships that are 
present in $G$, but cannot reliably detect the \emph{absence} of ancestral relationships.
We even allow our oracles to contradict themselves by giving responses that would lead to a 
cyclic graph when combined. If we use such oracles for DAG construction, we will
introduce superfluous edges into our graph.

Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$, and $0$ otherwise. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \citep{Spirtes2001}.

We now define two core procedures that use ancestral and $d$-separation queries
to iteratively change a current DAG structure. The procedure 
\textsc{Expand} (Algorithm~\ref{algo:expand}) uses CI  
information to search for unexplained associations in the graph and 
then uses domain knowledge to determine ancestry. The 
procedure \textsc{Prune} (Algorithm~\ref{algo:prune}) uses
CI information to remove superfluous edges from the
graph.

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    $L \gets L \cup \{ e \} $ \;
	    \If{ $ (V, E \cup L) $ is cyclic }{
		    $ L' \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}_G) $ \;
		    $ B \gets B \cup (L \setminus L') $ \;
		    $ L \gets L'$
	    }
	    % \If{ $(V,E \cup L \cup e)$ is acyclic }{
            % 	$L \gets L \cup  \{ e \} $\;
	    % } \Else {
            % 	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    % }
        }
	\If{$|L|\geq k$}{\KwGoTo 18\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E,B)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycle}{FixCycle}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D}_G $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ x_i \rightarrow x_j $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}_G(x_i, x_j)=1$
}{
			$ R \gets R \cup \{ x_i \rightarrow x_j \} $ \;

		}
	}
	\Return $ L \setminus R $ \;
}
\caption{Fixing cycles by removing incorrect edges.}
\label{algo:fix_cycles}
\end{algorithm}
